---
title: "Hands-on Exercise 8"
date: "March 4, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
fig_width: 15
fig_height: 10
---

# `Modelling, Visualising and Analysing Network Data with R`

## **1 Getting Started**

### **1.1 Installing and loading the packages**

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts)
```

### **1.2 Importing Data into R**

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

Examine the structure of the data frame using *glimpse()* of **dplyr**.

```{r}
glimpse(GAStech_edges)
```

The code chunk below will be used to perform the changes.

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

Table below shows the data structure of the reformatted *GAStech_edges* data frame

```{r}
glimpse(GAStech_edges)
```

A close examination of *GAStech_edges* data.frame reveals that it consists of individual e-mail flow records. This is not very useful for visualisation.

In view of this, we will aggregate the individual by date, senders, receivers, main subject and day of the week.

The code chunk:

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

```{r}
glimpse(GAStech_edges_aggregated)
```

## **2 Creating network objects using tidygraph**

### **2.1 The tbl_graph object**

Two functions of **tidygraph** package can be used to create network objects:

-   [`tbl_graph()`](https://tidygraph.data-imaginist.com/reference/tbl_graph.html) creates a **tbl_graph** network object from nodes and edges data.

-   [`as_tbl_graph()`](https://tidygraph.data-imaginist.com/reference/tbl_graph.html) converts network data and objects to a **tbl_graph** network. Below are network data and objects supported by `as_tbl_graph()`

    -   a node data.frame and an edge data.frame,

    -   data.frame, list, matrix from base,

    -   igraph from igraph,

    -   network from network,

    -   dendrogram and hclust from stats,

    -   Node from data.tree,

    -   phylo and evonet from ape, and

    -   graphNEL, graphAM, graphBAM from graph (in Bioconductor).

### **2.2 The dplyr verbs in tidygraph**

*activate()* verb from **tidygraph** serves as a switch between tibbles for nodes and edges. All dplyr verbs applied to **tbl_graph** object are applied to the active tibble.

### **2.3 Using `tbl_graph()` to build tidygraph data model**

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

### **2.4 Reviewing the output tidygraph’s graph object**

```{r}
GAStech_graph
```

### **2.5 Reviewing the output tidygraph’s graph object**

-   The output above reveals that *GAStech_graph* is a tbl_graph object with 54 nodes and 4541 edges.

-   The command also prints the first six rows of “Node Data” and the first three of “Edge Data”.

-   It states that the Node Data is **active**. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time

### **2.6 Changing the active object**

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## **3 Plotting Static Network Graphs with ggraph package**

### **3.1 Plotting a basic network graph**

The code chunk below uses [*ggraph()*](https://ggraph.data-imaginist.com/reference/ggraph.html), [*geom-edge_link()*](https://ggraph.data-imaginist.com/reference/geom_edge_link.html) and [*geom_node_point()*](https://ggraph.data-imaginist.com/reference/geom_node_point.html) to plot a network graph by using *GAStech_graph*. Before your get started, it is advisable to read their respective reference guide at least once.

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

### **3.2 Changing the default network graph theme**

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### **3.3 Changing the coloring of the plot**

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

### **3.4 Fruchterman and Reingold layout**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### **3.5 Modifying network nodes**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

### **3.6 Modifying edges**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

## **4 Creating facet graphs**

### **4.1 Working with *facet_edges()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

### **4.2 Working with *facet_edges()***

The code chunk below uses *theme()* to change the position of the legend.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

### **4.3 A framed facet graph**

The code chunk below adds frame to each graph.

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

### **4.4 Working with *facet_nodes()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## **5 Network Metrics Analysis**

### **5.1 Computing centrality indices**

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

### **5.2 Visualising network metrics**

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

### **5.3 Visualising Community**

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  

g + theme_graph()
```

## **6 Building Interactive Network Graph with visNetwork**

### **6.1 Data preparation**

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

### **6.2 Plotting the first interactive network graph with layout**

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

### **6.3 Working with visual attributes - Nodes**

visNetwork() looks for a field called “group” in the nodes object and colour the nodes according to the values of the group field.

The code chunk below rename Department field to group.

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

When we rerun the code chunk below, visNetwork shades the nodes by assigning unique colour to each category in the *group* field.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### **6.4 Working with visual attributes - Edges**

In the code run below *visEdges()* is used to symbolise the edges.\
- The argument *arrows* is used to define where to place the arrow.\
- The *smooth* argument is used to plot the edges using a smooth curve.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### **6.5 Interactivity**

In the code chunk below, *visOptions()* is used to incorporate interactivity features in the data visualisation.

-   The argument *highlightNearest* highlights nearest when clicking a node.

-   The argument *nodesIdSelection* adds an id node selection creating an HTML select element.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
