---
title: "2 Beyond ggplot2 Fundamentals"
author: "Zhao Xinyue"
date: "January 17, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

# 1.Getting Started

## 1.1 Install and load the required R libraries

The name and function of the new packages that will be used for this exercise are as follow:

-   ggrepel: an R package provides a way to prevent labels from overlapping in ggplot2 plots.

-   ggthemes: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.

-   hrbrthemes: an R package provides another set of visually appealing themes and formatting options for ggplot2.

-   patchwork: an R package provides a way to combine multiple ggplot2 plots into a single figure.

```{r}
# eval: false
pacman::p_load(tidyverse, patchwork, 
               ggthemes, hrbrthemes,
               ggrepel)
```

## 1.2 Import the data

```{r}
# eval: false
exam_data <- read_csv("data/Exam_data.csv")
```

# 2.Exercises

## 2.1 Working with ggrepel package

ggrepel is an extension of ggplot2 package which provides geoms for ggplot2 to repel overlapping text as in our examples on the right.

```{r}
# eval: false
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold") +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## 2.2 Working with themes

8 Built-in themes: theme_gray(), theme_bw(), theme_classic(), theme_dark(), theme_light(), theme_linedraw(), theme_minimal(), and theme_void(). Refer to [**here**](https://ggplot2.tidyverse.org/reference/#themes)

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_gray() +
  ggtitle("Ditribution of Maths scores")
```

### 2.2.1 Working with ggthems package

In the example below, The Economist theme is used.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_economist() +
  ggtitle("Ditribution of Maths scores")
```

### 2.2.2 Working with hrbthems package

[**hrbrthemes**](https://cinc.rud.is/web/packages/hrbrthemes/)This focuses more on the typographic elements, labels and fonts.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

The second goal centers around productivity for a production workflow. In fact, this “production workflow” is the context for where the elements of hrbrthemes should be used.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_ipsum(axis_title_size = 18,          #increase font size of the axis title to 18
              base_size = 15,                #increase the default axis label to 15
              grid = "Y") +                  #only keep the y-axis grid line -> remove the x-axis grid lines
  ggtitle("Ditribution of Maths scores")
```

## 2.3 Working with patchwork

Creating composite plot by combining multiple graphs. Start with creating three statistical graphics below.

```{r}
#creating histogram
p1 <- ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) +
  ggtitle("Ditribution of Maths scores")

p2 <- ggplot(data = exam_data,
       aes(x = ENGLISH)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) +
  ggtitle("Ditribution of English scores")

#creating scatterplot
p3 <- ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

Creating patchwork.(Refer [**here**](https://patchwork.data-imaginist.com/articles/guides/assembly.html) for more details)

-   Use ‘+’ sign to create two columns layout

-   Use ‘/’ sign to create two row layout (stack)

-   Use ‘()’ sign to create subplot group

-   Use ‘\|’ sign to place the plots besisde each other

**Example 1**

```{r}
p1 + p2 / p3
```

**Example 2**

```{r}
(p1 / p2) | p3
```

**Example 3: It will auto-tag the subplots in text**

```{r}
((p1 / p2) | p3) +
  plot_annotation(tag_levels = 'I')
```

**Example 4: Combine patchwork and themes**

```{r}
((p1 / p2) | p3) & theme_economist()
```

**Example 5: Insert another plot in a plot [using insert_element()](https://patchwork.data-imaginist.com/reference/inset_element.html)**

```{r}
p3 + inset_element(p2,
                   left = 0.02,
                   bottom = 0.7,
                   right = 0.5,
                   top = 1)
```
