---
title: "Take-home_Ex01"
date: "January 20, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

1.OVERVIEW

Despite official claims of educational equity in Singapore, public perception highlights persistent disparities based on socioeconomic status, school type, and family background.

2.OBJECTIVE

Using appropriate **Exploratory Data Analysis (EDA)** methods and **ggplot2** functions to get insight.

-   The distribution of Singapore students’ performance in mathematics, reading, and science.

-   The relationship between these performances with schools, gender and socioeconomic status of the students.

3.DATA PREPARATION

3.1 Loading R packages

```{r}
pacman::p_load(tidyverse,haven)
```

3.2 Data set

As we are foucesd on Singapore database. So firstly we should filter Singapore dataset. Refer to [In-class Exercise 1](https://isss608-ay2023-24t2.netlify.app/in-class_ex/in-class_ex01/in-class_ex01) for more details.

```{r}
# eval: false
stu_qqq_SG <-
  read_rds("data/stu_qqq_SG.rds")
```

3.3 Data Preparation
```{r}
# eval: false
grep("PV", colnames(stu_qqq_SG), value = TRUE)
```

```{r}
# eval: false
if (!require(devtools)) {
  install.packages("devtools")
  library(devtools)
}
devtools::install_github("eldafani/intsvy")
```

```{r}
# eval: false
pacman::p_load(intsvy)
```

```{r}
# eval: false
pisa_data <- readRDS("data/stu_qqq_SG.rds")
```

loading package
```{r}
pacman::p_load(dplyr)
```

```{r}
any(is.na(pisa_data[, paste0("PV", 1:10, "MATH")]))
```

```{r}
# eval: false
math_mean <- with(data = pisa_data, expr = {
  mean(c(PV1MATH, PV2MATH, PV3MATH, PV4MATH, PV5MATH,
         PV6MATH, PV7MATH, PV8MATH, PV9MATH, PV10MATH), na.rm = TRUE)
})
```

::: {.panel-tabset}
## plot1
```{r}
library(ggplot2)

ggplot(pisa_data, aes(x = PV1MATH)) +
  geom_histogram(binwidth = 10, fill = "blue") +
  theme_minimal() +
  ggtitle("Distribution of  PV1Math Scores")
```
## plot2
```{r}
library(ggplot2)

ggplot(pisa_data, aes(x = PV2MATH)) +
  geom_histogram(binwidth = 10, fill = "blue") +
  theme_minimal() +
  ggtitle("Distribution of  PV2Math Scores")
```
:::
