---
title: "Take-home_Ex01"
date: "January 20, 2024"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

## 1.OVERVIEW

Despite official claims of educational equity in Singapore, public perception highlights persistent disparities based on socioeconomic status, school type, and family background.

## 2.OBJECTIVE

Using appropriate **Exploratory Data Analysis (EDA)** methods and **ggplot2** functions to get insight.

-   The distribution of Singapore students’ performance in mathematics, reading, and science.

-   The relationship between these performances with schools, gender and socioeconomic status of the students.

## 3.DATA PREPARATION

### 3.1 Loading R packages

```{r}
pacman::p_load(tidyverse,haven,dplyr,tidyr,ggplot2,patchwork,ggthemes)
```

### 3.2 Data set

As we are focursed on Singapore database. So firstly we should filter Singapore dataset. Refer to [In-class Exercise 1](https://isss608-ay2023-24t2.netlify.app/in-class_ex/in-class_ex01/in-class_ex01) for more details.

```{r}
# eval: false
stu_qqq_SG <-
  read_rds("data/stu_qqq_SG.rds")
```

### 3.3 Data Preparation

#### 1.Find the relative math score data in 'stu_qqq_SG' dataset

```{r}
# eval: false
# Select the specific columns and view the first few rows
head(select(stu_qqq_SG, PV1MATH, PV2MATH, PV3MATH, PV4MATH, PV5MATH, PV6MATH, PV7MATH, PV8MATH, PV9MATH, PV10MATH))
```

Similarly, we can find the reading and science score data and then get the average value for each course.

#### 2.Add new columns for math,reading and science

Calculate the average score of the three courses separately.

```{r}
stu_qqq_SG <- stu_qqq_SG %>%
  mutate(
    Math = rowMeans(select(., starts_with("PV1MATH"), ends_with("PV10MATH")), na.rm = TRUE),
    Reading = rowMeans(select(., starts_with("PV1READ"), ends_with("PV10READ")), na.rm = TRUE),
    Science = rowMeans(select(., starts_with("PV1SCIE"), ends_with("PV10SCIE")), na.rm = TRUE)
  )
```

#### 3.Create new clean table for later-on analysis

Select relative columns and create new_data with 7 columns and 6606 ids. Then see the total view of the data, where find that there are 47 missing values which can be considered deleted.

```{r}
new_data <- select(stu_qqq_SG,CNTSTUID,CNTSCHID,ST004D01T, ESCS, Math, Reading, Science)
summary(new_data)
new_data <- new_data %>% filter(complete.cases(.))
```

## 4.DATA ANALYSIS

### 4.1 The distribution of Singapore students’ performance in each course

```{r}
long_data <- pivot_longer(new_data, 
                          cols = c("Math", "Reading", "Science"),
                          names_to = "subject", 
                          values_to = "score")
ggplot(long_data, aes(x = subject, y = score, fill = subject)) +
  geom_boxplot() +  
  scale_fill_manual(values = c("Math" = "magenta", "Reading" = "green", "Science" = "yellow")) + # 为不同的科目设置颜色
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red") +
  
  theme_minimal() +
  labs(title = "Distribution of Scores in Math, Reading, and Science",
       x = "Subject",
       y = "Average Score") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

| ⛳
| From the plot, we could see that for students performance in three subjects,math is the most highest among three courses, and the next up higher is science. Reading the lowest.

### 4.2 Subject Performance by gender

```{r}
ggplot(long_data, aes(x = subject, y = score, fill = factor(ST004D01T, labels = c("Female", "Male")))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  labs(title = "Performance Distribution by Subject and Gender",
       x = "Subject",
       y = "Score",
       fill = "Gender") +
  theme_minimal() +
  theme(legend.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

| ⛳ 
| From the boxplot, we could see that male tends to get higher scores in math and science than female. Female tends to perform better in Reading.

### 4.3 Subject Performance by different schools

| ⛳
| From the plot, we could see that students in school 003 get highest score in math and science. Students in school 062 get the highest score in subject. However, students in school 149 perform the worst in all three courses.

```{r}
# Calculate the average scores for each school and subject
school_averages <- new_data %>%
  group_by(CNTSCHID) %>%
  summarize(
    Math = mean(Math, na.rm = TRUE),
    Reading = mean(Reading, na.rm = TRUE),
    Science = mean(Science, na.rm = TRUE),
    .groups = 'drop'
  )

# Reshape the data to a long format for plotting
school_averages_long <- school_averages %>%
  pivot_longer(
    cols = c(Math, Reading, Science),
    names_to = "Subject",
    values_to = "Average_Score"
  )

# Find the top and bottom schools for each subject
top_schools <- school_averages_long %>%
  group_by(Subject) %>%
  slice_max(Average_Score, n = 1) %>%
  ungroup()

bottom_schools <- school_averages_long %>%
  group_by(Subject) %>%
  slice_min(Average_Score, n = 1) %>%
  ungroup()

# Create the scatter plot
ggplot(school_averages_long, aes(x = Subject, y = Average_Score)) +
  geom_point(aes(color = CNTSCHID), position = position_jitterdodge()) +
  labs(title = "Subject Performance by Different Schools",
       x = "Subject", 
       y = "Average Score") +
  theme_minimal() +
  geom_text(data = top_schools, aes(label = CNTSCHID), vjust = 2, color = "blue") +
  geom_text(data = bottom_schools, aes(label = CNTSCHID), vjust = 1, color = "red")
```

### 4.4 Subject Performance by socioeconomic status

```{r}
theme_setting <- theme(
  plot.title = element_text(size = 9),
  plot.subtitle = element_text(size = 8),
  plot.caption = element_text(size = 8),
  axis.text.x = element_text(angle = 45, hjust = 1) # Adjust for better readability
)

# Adjust the alpha and size in geom_point for transparency and smaller points
p1 <- ggplot(data=new_data, aes(x= ESCS, y=Math, color = "Math")) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method=lm, linewidth=0.5) +  
  coord_cartesian(xlim=c(-3,3), ylim=c(250,850)) +
  ggtitle("Math vs. ESCS") +  
  theme_setting +
  scale_color_manual(values = c("Math" = "magenta"))

p2 <- ggplot(data=new_data, aes(x= ESCS, y=Reading, color = "Reading")) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method=lm, linewidth=0.5) +  
  coord_cartesian(xlim=c(-3,3), ylim=c(250,850)) +
  ggtitle("Reading vs. ESCS") +
  theme_setting +
  scale_color_manual(values = c("Reading" = "green"))

p3 <- ggplot(data=new_data, aes(x= ESCS, y=Science, color = "Science")) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method=lm, linewidth=0.5) +  
  coord_cartesian(xlim=c(-3,3), ylim=c(250,850)) +
  ggtitle("Science vs. ESCS") +
  theme_setting +
  scale_color_manual(values = c("Science" = "yellow"))


patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

| ⛳
| From the plot, we could see that all three subjects—Math, Reading, and Science—show a positive correlation with the Economic, Social, and Cultural Status (ESCS) index. Most of the students are in the middle class.

## 5.CONCLUSION

\
Students generally score better in Math compared to Reading. Girls typically excel in Reading but may not perform as well in Science. Boys often do better in Math and Science.

School 003 leads with top scores in Math and Science, which may reflect its excellent programs in these fields. School 062 shines in Reading, hinting at a strong focus on literacy. Meanwhile, school 149 lags in all subjects, which could signal various difficulties at the school.

Higher socioeconomic status is linked to better grades for students.
