---
title: "Take-home_Ex01"
date: "January 18, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

1.OVERVIEW

Despite official claims of educational equity in Singapore, public perception highlights persistent disparities based on socioeconomic status, school type, and family background.

2.OBJECTIVE

Using appropriate **Exploratory Data Analysis (EDA)** methods and **ggplot2** functions to get insight.

-   The distribution of Singapore students’ performance in mathematics, reading, and science.

-   The relationship between these performances with schools, gender and socioeconomic status of the students.

3.DATA PREPARATION

3.1 Loading R packages

```{r}
pacman::p_load(tidyverse,haven)
```

3.2 Data set

As we are foucesd on Singapore database. So firstly we should filter Singapore dataset. Refer to [In-class Exercise 1](https://isss608-ay2023-24t2.netlify.app/in-class_ex/in-class_ex01/in-class_ex01) for more details

```{r}
# eval: false
stu_qqq_SG <-
  read_rds("data/stu_qqq_SG.rds")
```

3.3 Data Preparation

```{r}
grep("PV", colnames(stu_qqq_SG), value = TRUE)
```
